Challenging Fuel Cycle Modeling Assumptions: Facility and Time Step Discretization Effects

### 2015-10-20 paper.tex
add github PR comments for provenance
spell check

\myabstract{
Due to the diversity of fuel cycle simulator modeling assumptions, direct
comparison and benchmarking can be difficult.  In 2012 the Organisation for
Economic Co-operation and Development (OECD) completed a benchmark study that
is perhaps one of the most complete published comparisons performed.  Despite
this, however, various results from the simulators were often significantly
different due to inconsistency in modeling decisions involving reprocessing
strategies, refueling behavior, reactor end-of-life handling, etc.  This work
attempts to identify and quantify the effects of selected modeling choices
that may sometimes be taken for granted in the fuel cycle modeling domain.
Four scenarios are compared using combinations of either fleet-based or
individual reactors with either monthly or quarterly (3-month) time steps.
The Cyclus fuel cycle simulator's plug-in facility cability along with its
market-like dynamic material routing allow it to be used as a level playing
field for comparing the scenarios.  When under supply-constraint pressure, the
four cases exhibit noticably different behavior.  Fleet-based modeling is more
efficient in supply-constrained environments at the expense of losing insight
on issues such as realistically suboptimal fuel distribution and challenges in
reactor refueling cycle staggering.  Finer-grained time steps also enable more
efficient material use in supply-constrained environments.  Large simulations
with fleet-based reactors run much more quickly than their individual reactor
counterparts.
}

@@

The diversity of assumptions embedded within various fuel cycle simulators pose challenges for direct
comparison and benchmarking.  In 2012 the Organisation for
Economic Co-operation and Development (OECD) completed a benchmark study
\cite{oecd2012benchmark} that is perhaps one of the most complete published
comparisons performed.  Despite this, however, various results from the
different simulators were often significantly different due to inconsistent
modeling decisions involving reprocessing strategies, refueling behavior,
reactor end-of-life handling, etc.  This work attempts to identify and
quantify the effects of selected modeling choices relative to facility modeling
discretization (i.e. individual facilities vs. aggregate fleets) and time step
duration.  A better understanding of how these affect fuel cycle dynamics can
allow for making more deliberate decisions with respect to tradeoffs such as
computational investment vs. realism.

@@

One of the major motivations for the development of the Cyclus fuel cycle
simulator \cite{cyclus_2015} was creating a simulation environment that is
more expressive than more traditional fuel cycle simulators.  Cyclus was
designed with a plugin-style architecture that allows different models for
facility types, known as \emph{archetypes}, to be swapped in and out easily.

As a Cyclus simulation walks through each time step, facilities engage in a
market-like dynamic resource exchange (DRE).  They broadcast requests for
material and potential supplying facilities respond with bids. The requesting facilities then
express preferences between all their bids, and the DRE then generates a network
flow problem that is optimized to resolve which transactions actually occur.
The plugin architecture combined with the DRE creates a powerful and flexible
environment for the natural comparison of different modeling paradigms.
Cyclus' module plug-in based architecture in concert with its DRE has the ability to
model fleet-based facilities.  Additionally, however, Cyclus can model
facilities individually with discrete flows enabling the investigation of
interesting real-world effects such as competition, reactor cycle staggering,
and individual reactor outages.

Cyclus was designed to operate natively in unix/linux type environments with
flexible and scriptable usage and input file format.  Cyclus stores its
comprehensive output data in a powerful open-source database format.  In addition
to information about material flows and facility deployments, Cyclus'
single-file database contains the simulation input file, version information
about the archetypes used in a simulation, the version of Cyclus used to run
the simulation, and the installed versions of Cyclus dependencies. This and
other features of Cyclus lend themselves well to automation, research
provenance and reproducibility, and large-scale computing.

@@

A fleet reactor archetype was created for Cyclus.  Although the modeling of
individual, discrete facilities was a motivation for Cyclus, its flexibility
makes it relatively straightforward to create plug-in archetypes to model
anywhere along the spectrum from individual, discrete facilities to
continuous flow fleet facilities.  In order to contrast well with the individual,
single-facility granularity modeling that is part of the standard Cyclus archetypes, 
a fleet archetype was created for this work with aggregate facilitiy behavior and
continuous flow.

The fleet reactor, although a single entity modeling many reactors, is
actually designed to look like many individual reactors to the cyclus
simulator kernel. This "trick" allows other plugins for managing facility
deployments to transparently work with the fleet reactor without having to
know anything about the fleet paradigm.  The fleet reactor has several
important characteristics described below:

@@

    \item Generated power at any timestep, $P_G$, is based on the number of
      operating reactors:

@@

One characteristic of this fleet reactor implementation is that it has perfect
fuel sharing and perfect cycle staggering among the individual reactors it
models.  The first discharge of spent fuel begins immediately on
the first time step of operation rather than after a complete refueling cycle.
Fuel shortages also only cause proportional loss of capacity rather than
whole-reactor quantized shutdown.  The full implementation of the fleet-based
reactor is publicy available for download and use \cite{Carlsen2015}.

@@

The chosen scenario approximates a transition from a the current once-through light water
reactor (LWR) fleet to a full sodium fast reactor (SFR) fuel cycle within evaluation group 23 (EG-23)
in the DOE's Advanced Nuclear Fuel Cycle Options report
\cite{wigeland_nuclear_2014}. Scenario and transition details were patterned
after ongoing work by the DOE in their Fuel Cycle Options campaign.
%% Not sure if we should make such an explicit reference to the ongoing FCO
%% transition work until they publish something. On one hand, it reveals
%% things they are secretive about. On the other hand, we should give them due
%% credit.

@@

Instead of annual deployments, a longer 21-month deployment period is used
because it provides better natural staggering for the reactor refueling
cycles. The same deployment schedule is used for all four cases; power
capacity is built along a curve starting at 100GWe in year zero following the
1\% exponential growth curve.  This particular deployment schedule is chosen
because it causes significant fuel shortages providing interesting features to
investigate and compare.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=1.0\columnwidth]{exp2/sync-cycle.eps}
    \caption[The cycle staggering effect]{
        In this scenario, reactors are deployed annualy with a refueling cycle
        length (including outage) of 18 months.  So all reactors deployed
        every 3rd year refuel together.  All initial 100 LWRs start out with
        their cycles synced as well.
    }
    \label{fig:sync-cycle}
\end{figure}

@@

There are also performance differences between the two modeling paradigms.  A
single 200 year simulation with several hundred fleet-based reactors takes
about 2 seconds to run on a typical desktop machine where the same simulation
using individual reactor facilities takes about 20 seconds.  This order of
magnitude difference is an important tradeoff to keep in mind when doing fuel
cycle analysis in general, particularly when considering optimization or
sensitivity studies where running thousands or millions of simulations may be
desirable.  The time step duration has less impact on performance because the
amount of work being done in the simulation is similar; in the monthly time
step cases most facilities (i.e. all reactors) only perform actions every
refueling cycle rather than every time step.

@@

During the initial 100 years, cases MI and QI have larger variance in power
output than cases MF and QF.  This is caused by minor refueling cycle
synchronization.  The \emph{cycle staggering effect} becomes much more visible
during and after the fuel shortage as evidenced by the larger swings in power
generation from time step to time step.  Initially, the reactor cycles are
staggered well.  During the shortage, several reactors that previously had
staggered cycles are all offline together waiting for fuel.  On time steps
where reactors retire (not just during the shortage), new deployments are made
to replace them in addition to new deployments made to address power demand
growth.  Because the initial reactors retire in waves somewhat close together,
there are corresponding waves of deployment, and these waves are echoed every
80 years (the reactor lifetime).  These waves of deployment cause surges in
recycled fuel availability when they begin to discharge their fuel
%% these need to be LWR waves, right, or they'd just make things worse?
that cause many of the reactors offline during the
shortage to receive fuel and come online together.  The net effect is that the
fuel shortage degrades reactor cycle stagering overall.  As the simulation
continues, however, new reactors continue to be deployed and reactors with
synchornized cycles retire resulting in a gradual return to better
staggering visible by the end of the 200 years.

\begin{figure*}[!h]
    \centering
    \includegraphics[width=1.0\textwidth]{exp2/power.eps}
    \caption[Generated power]{
        Generated power for all four cases.  A fuel shortage occurs from about
        year 100 to about year 140.  In cases MI and QI with individual reactor
        modeling, refueling cycles become much more synchronized during and
        after the fuel shortage resuting in large power fluctuations - an artifact of the
        \emph{cycle staggering effect}.
    }
    \label{fig:power}
\end{figure*}

\begin{figure*}[!h]
    \centering
    \includegraphics[width=1.0\textwidth]{exp2/power-rel.eps}
    \caption[Normalized power]{
        Generated power is normalized to the expected 1\% exponential growth
        curve for all four cases.  A fuel shortage occurs from about year 100
        to about year 140.  The \emph{cycle staggering effect} can be seen as
        excessive divergence above and below $1.0$ in cases MI and QI where
        the shortage increases cycle synchronization significantly. The
        \emph{quantized shutdown effect} can also be seen in the discrepency
        between quarterly and monthly time steps for fleet reactor modeling
        (cases MF and QF) during the shortage; a longer time step causes more
        reactor outage than necessary. The consistent jumping up and down for
        the fleet reactor cases (even pre-shortage) is due to the mis-alignment of reactors going
        offline (multiples of 12 months) and the build period (21 months).
    }
    \label{fig:power-rel}
\end{figure*}

@@

%% What about a different measure of fuel sharing effect?  Perfect sharing
%% in a fleet reactor results in N_o reactors operating.  For individual
%% reactors, that would mean floor(N_o) reactors.  Can't we measure the
%% actual number of reactors that are operating, N_actual, and accumulate
%% floor(N_o) - N_actual as a measure of the effect of fuel sharing?  Would
%% this be substantially different?

One useful over-estimate of the fuel sharing inefficiency is to assume every
one of the "wasted" batches could have been given to a reactor enabling it to
operate.  Multiplying each of these cumulative batches by the fast reactor power capacity
(i.e. 450 MWe) provides a way to compare fuel sharing inefficiency with the
overall fuel outage.  This results in a cumulative energy deficit caused by
poor fuel sharing of roughly 2,000 $\text{GWe} \cdot \text{months}$ and 6,600
$\text{GWe} \cdot \text{months}$ for cases MI and QI respectively.  These
approximations account for a significant portion of the difference between individual
and fleet cases in Figure \ref{fig:unfueled} suggesting that most of the
discrepency in cumulative outage between fleet and individual modeling is
caused by inefficient fuel sharing in the individual reactor cases.

@@

Throughout most of the shortage, the aggregate outflow of Pu from separated
inventory is roughly the same in cases MF and QF.  However, due to the longer
time steps in case QF, the shortage starts slightly sooner because Pu inventory
must be maintained at a higher limit as described above.  The small difference
in outage start time has feed-back effects.  Not only are extra reactors
offline, but those extra reactors are also discharging less aggregate spent
fuel for recycling.  This increases the discrepency between the two
cases. Eventually the shortage in case MF ends slightly before it does in case
QF (shown most clearly in Figure \ref{fig:puinv-compare-zoom}). This feedback effect occurs because the reactors have a
conversion ratio greater than 1.0.

@@

The value of realism in fuel cycle modeling is greatly be enhanced by an
associated understanding of how outcomes are affected by added fidelity.
Cyclus' flexibility for accomodating different modeling choices uniquely
enables many interesting comparisons.  Continued work in this direction has
great potential to to improve the integrity of fuel cycle analysis generally.

>>>>

\myabstract{
Due to the diversity of fuel cycle simulator modeling assumptions, direct
comparison and benchmarking can be difficult.  In 2012 the Organisation for
Economic Co-operation and Development (OECD) completed a benchmark study that
is perhaps one of the most complete published comparisons performed.  Despite
this, however, various results from the simulators were often significantly
different due to inconsistency in modeling decisions involving reprocessing
strategies, refueling behavior, reactor end-of-life handling, etc.  This work
attempts to identify and quantify the effects of selected modeling choices
that may sometimes be taken for granted in the fuel cycle modeling domain.
Four scenarios are compared using combinations of either fleet-based or
individual reactors with either monthly or quarterly (3-month) time steps.
The Cyclus fuel cycle simulator's plug-in facility capability along with its
market-like dynamic material routing allow it to be used as a level playing
field for comparing the scenarios.  When under supply-constraint pressure, the
four cases exhibit noticeably different behavior.  Fleet-based modeling is more
efficient in supply-constrained environments at the expense of losing insight
on issues such as realistically suboptimal fuel distribution and challenges in
reactor refueling cycle staggering.  Finer-grained time steps also enable more
efficient material use in supply-constrained environments.  Large simulations
with fleet-based reactors run much more quickly than their individual reactor
counterparts.
}

@@

The diversity of assumptions embedded within various fuel cycle simulators pose challenges for direct
comparison and benchmarking.  In 2012 the Organisation for
Economic Co-operation and Development (OECD) completed a benchmark study
\cite{oecd2012benchmark} that is perhaps one of the most complete published
comparisons performed.  Despite this, however, various results from the
different simulators were often significantly different due to inconsistent
modeling decisions involving reprocessing strategies, refueling behavior,
reactor end-of-life handling, etc.  This work attempts to identify and
quantify the effects of selected modeling choices relative to facility modeling
discretization (i.e. individual facilities vs. aggregate fleets) and time step
duration.  A better understanding of how these affect fuel cycle dynamics can
allow for making more deliberate decisions with respect to trade-offs such as
computational investment vs. realism.

@@

One of the major motivations for the development of the Cyclus fuel cycle
simulator \cite{cyclus_2015} was creating a simulation environment that is
more expressive than more traditional fuel cycle simulators.  Cyclus was
designed with a plug-in style architecture that allows different models for
facility types, known as \emph{archetypes}, to be swapped in and out easily.

As a Cyclus simulation walks through each time step, facilities engage in a
market-like dynamic resource exchange (DRE).  They broadcast requests for
material and potential supplying facilities respond with bids. The requesting facilities then
express preferences between all their bids, and the DRE then generates a network
flow problem that is optimized to resolve which transactions actually occur.
The plug-in architecture combined with the DRE creates a powerful and flexible
environment for the natural comparison of different modeling paradigms.
Cyclus' module plug-in based architecture in concert with its DRE has the ability to
model fleet-based facilities.  Additionally, however, Cyclus can model
facilities individually with discrete flows enabling the investigation of
interesting real-world effects such as competition, reactor cycle staggering,
and individual reactor outages.

Cyclus was designed to operate natively in Unix/Linux type environments with
flexible and scriptable usage and input file format.  Cyclus stores its
comprehensive output data in a powerful open-source database format.  In addition
to information about material flows and facility deployments, Cyclus'
single-file database contains the simulation input file, version information
about the archetypes used in a simulation, the version of Cyclus used to run
the simulation, and the installed versions of Cyclus dependencies. This and
other features of Cyclus lend themselves well to automation, research
provenance and reproducibility, and large-scale computing.

@@

A fleet reactor archetype was created for Cyclus.  Although the modeling of
individual, discrete facilities was a motivation for Cyclus, its flexibility
makes it relatively straightforward to create plug-in archetypes to model
anywhere along the spectrum from individual, discrete facilities to
continuous flow fleet facilities.  In order to contrast well with the individual,
single-facility granularity modeling that is part of the standard Cyclus archetypes, 
a fleet archetype was created for this work with aggregate facility behavior and
continuous flow.

The fleet reactor, although a single entity modeling many reactors, is
actually designed to look like many individual reactors to the Cyclus
simulator kernel. This "trick" allows other plug-ins for managing facility
deployments to transparently work with the fleet reactor without having to
know anything about the fleet paradigm.  The fleet reactor has several
important characteristics described below:

@@

    \item Generated power at any time step, $P_G$, is based on the number of
      operating reactors:

@@

One characteristic of this fleet reactor implementation is that it has perfect
fuel sharing and perfect cycle staggering among the individual reactors it
models.  The first discharge of spent fuel begins immediately on
the first time step of operation rather than after a complete refueling cycle.
Fuel shortages also only cause proportional loss of capacity rather than
whole-reactor quantized shutdown.  The full implementation of the fleet-based
reactor is publicly available for download and use \cite{Carlsen2015}.

@@

The chosen scenario approximates a transition from the current once-through light water
reactor (LWR) fleet to a full sodium fast reactor (SFR) fuel cycle within evaluation group 23 (EG-23)
in the DOE's Advanced Nuclear Fuel Cycle Options report
\cite{wigeland_nuclear_2014}. Scenario and transition details were patterned
after ongoing work by the DOE in their Fuel Cycle Options campaign.
%% Not sure if we should make such an explicit reference to the ongoing FCO
%% transition work until they publish something. On one hand, it reveals
%% things they are secretive about. On the other hand, we should give them due
%% credit.

@@

Instead of annual deployments, a longer 21-month deployment period is used
because it provides better natural staggering for the reactor refueling
cycles. The same deployment schedule is used for all four cases; power
capacity is built along a curve starting at 100 GWe in year zero following the
1\% exponential growth curve.  This particular deployment schedule is chosen
because it causes significant fuel shortages providing interesting features to
investigate and compare.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=1.0\columnwidth]{exp2/sync-cycle.eps}
    \caption[The cycle staggering effect]{
        In this scenario, reactors are deployed annually with a refueling cycle
        length (including outage) of 18 months.  So all reactors deployed
        every 3rd year refuel together.  All initial 100 LWRs start out with
        their cycles synced as well.
    }
    \label{fig:sync-cycle}
\end{figure}

@@

There are also performance differences between the two modeling paradigms.  A
single 200 year simulation with several hundred fleet-based reactors takes
about 2 seconds to run on a typical desktop machine where the same simulation
using individual reactor facilities takes about 20 seconds.  This order of
magnitude difference is an important trade-off to keep in mind when doing fuel
cycle analysis in general, particularly when considering optimization or
sensitivity studies where running thousands or millions of simulations may be
desirable.  The time step duration has less impact on performance because the
amount of work being done in the simulation is similar; in the monthly time
step cases most facilities (i.e. all reactors) only perform actions every
refueling cycle rather than every time step.

@@

During the initial 100 years, cases MI and QI have larger variance in power
output than cases MF and QF.  This is caused by minor refueling cycle
synchronization.  The \emph{cycle staggering effect} becomes much more visible
during and after the fuel shortage as evidenced by the larger swings in power
generation from time step to time step.  Initially, the reactor cycles are
staggered well.  During the shortage, several reactors that previously had
staggered cycles are all offline together waiting for fuel.  On time steps
where reactors retire (not just during the shortage), new deployments are made
to replace them in addition to new deployments made to address power demand
growth.  Because the initial reactors retire in waves somewhat close together,
there are corresponding waves of deployment, and these waves are echoed every
80 years (the reactor lifetime).  These waves of deployment cause surges in
recycled fuel availability when they begin to discharge their fuel
%% these need to be LWR waves, right, or they'd just make things worse?
that cause many of the reactors offline during the
shortage to receive fuel and come online together.  The net effect is that the
fuel shortage degrades reactor cycle staggering overall.  As the simulation
continues, however, new reactors continue to be deployed and reactors with
synchronized cycles retire resulting in a gradual return to better
staggering visible by the end of the 200 years.

\begin{figure*}[!h]
    \centering
    \includegraphics[width=1.0\textwidth]{exp2/power.eps}
    \caption[Generated power]{
        Generated power for all four cases.  A fuel shortage occurs from about
        year 100 to about year 140.  In cases MI and QI with individual reactor
        modeling, refueling cycles become much more synchronized during and
        after the fuel shortage resulting in large power fluctuations - an artifact of the
        \emph{cycle staggering effect}.
    }
    \label{fig:power}
\end{figure*}

\begin{figure*}[!h]
    \centering
    \includegraphics[width=1.0\textwidth]{exp2/power-rel.eps}
    \caption[Normalized power]{
        Generated power is normalized to the expected 1\% exponential growth
        curve for all four cases.  A fuel shortage occurs from about year 100
        to about year 140.  The \emph{cycle staggering effect} can be seen as
        excessive divergence above and below $1.0$ in cases MI and QI where
        the shortage increases cycle synchronization significantly. The
        \emph{quantized shutdown effect} can also be seen in the discrepancy
        between quarterly and monthly time steps for fleet reactor modeling
        (cases MF and QF) during the shortage; a longer time step causes more
        reactor outage than necessary. The consistent jumping up and down for
        the fleet reactor cases (even pre-shortage) is due to the mis-alignment of reactors going
        offline (multiples of 12 months) and the build period (21 months).
    }
    \label{fig:power-rel}
\end{figure*}

@@

%% Paul Wilson says:
%%
%% What about a different measure of fuel sharing effect?  Perfect sharing
%% in a fleet reactor results in N_o reactors operating.  For individual
%% reactors, that would mean floor(N_o) reactors.  Can't we measure the
%% actual number of reactors that are operating, N_actual, and accumulate
%% floor(N_o) - N_actual as a measure of the effect of fuel sharing?  Would
%% this be substantially different?
%%
%% rwcarlsen responds:
%%
%% This could be interesting, but it is quantifying the significance of the
%% quantized shutdown effect discussed in the power production section of
%% results and doesn't belong here with the fuel sharing discussion - which is
%% only pseudo-related. Fuel sharing effects are when alternative sharing of
%% fuel between individual reactors could have resulted in different
%% online/offline capacity levels. These effects step on each other a tiny bit,
%% but I think it still makes much more sense to discuss/treat them separately.

One useful over-estimate of the fuel sharing inefficiency is to assume every
one of the "wasted" batches could have been given to a reactor enabling it to
operate.  Multiplying each of these cumulative batches by the fast reactor power capacity
(i.e. 450 MWe) provides a way to compare fuel sharing inefficiency with the
overall fuel outage.  This results in a cumulative energy deficit caused by
poor fuel sharing of roughly 2,000 $\text{GWe} \cdot \text{months}$ and 6,600
$\text{GWe} \cdot \text{months}$ for cases MI and QI respectively.  These
approximations account for a significant portion of the difference between individual
and fleet cases in Figure \ref{fig:unfueled} suggesting that most of the
discrepancy in cumulative outage between fleet and individual modeling is
caused by inefficient fuel sharing in the individual reactor cases.

@@

Throughout most of the shortage, the aggregate outflow of Pu from separated
inventory is roughly the same in cases MF and QF.  However, due to the longer
time steps in case QF, the shortage starts slightly sooner because Pu inventory
must be maintained at a higher limit as described above.  The small difference
in outage start time has feed-back effects.  Not only are extra reactors
offline, but those extra reactors are also discharging less aggregate spent
fuel for recycling.  This increases the discrepancy between the two
cases. Eventually the shortage in case MF ends slightly before it does in case
QF (shown most clearly in Figure \ref{fig:puinv-compare-zoom}). This feedback effect occurs because the reactors have a
conversion ratio greater than 1.0.

@@

The value of realism in fuel cycle modeling is greatly be enhanced by an
associated understanding of how outcomes are affected by added fidelity.
Cyclus' flexibility for accommodating different modeling choices uniquely
enables many interesting comparisons.  Continued work in this direction has
great potential to to improve the integrity of fuel cycle analysis generally.
